The Unicode Cookbook \linebreak for Linguists

### 2017-12-26 book/chapters/implementation.tex
closes issue #23
work towards issue 16 - wip

Note also that the performance with larger datasets differs, and the code is not
always as clean as we would like it to be. In sum, the two implementations
should be considered as `proof of concept' and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

@@

\begin{lstlisting}[language=bash]
  $ git clone https://github.com/bambooforest/segments.git
  $ cd segments
  $ python setup.py develop
\end{lstlisting}

@@

Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
help files with illustrative examples, and also a so-called `vignette' with
explanations and examples.

@@

Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) `shiny' webapps.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(`user-perceived characters', see Section~\ref{the-unicode-approach}). As it 
turns out, some of these graphemes are single codepoints, others are combinations
of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

It is important to realize that different ordering of the rules does not have
so-called `feeding' and `bleeding' effects as known from finite-state rewrite
rules.\footnote{`Bleeding' is the effect that the application of a rule changes
the string, so as to prevent a next rule to apply. `Feeding' is the opposite: a
specific rule will only be applied becuase a previous rule changed the string
already. The interaction of rules with such `feeding' and `bleeding' effects is
extremely difficult to predict.} The graphemic parsing advocated here works
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be `captured' in the 
string. All lines in the profile are processed in a specifed order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as `captured', but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the application of the `rules' is rather easy to predict.

>>>>

Note also that the performance with larger datasets differs, and the code is not
always as clean as we would like it to be. In sum, the two implementations
should be considered as proof of concept and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

@@

\begin{lstlisting}[language=bash]
  $ git clone https://github.com/cldf/segments
  $ cd segments
  $ python setup.py develop
\end{lstlisting}

@@

Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
help files with illustrative examples, and also a so-called vignette with
explanations and examples.

@@

Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) shiny webapps.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(called user-perceived characters, see Section~\ref{the-unicode-approach}). As it 
turns out, some of these graphemes are single codepoints, others are combinations
of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

It is important to realize that different ordering of the rules does not have
so-called feeding and bleeding effects as known from finite-state rewrite
rules.\footnote{Bleeding is the effect that the application of a rule changes
the string, so as to prevent a next rule to apply. Feeding is the opposite: a
specific rule will only be applied becuase a previous rule changed the string
already. The interaction of rules with such feeding and bleeding effects is
extremely difficult to predict.} The graphemic parsing advocated here works
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be captured in the 
string. All lines in the profile are processed in a specifed order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as captured, but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the application of the rules' is rather easy to predict.

### 2017-12-26 book/chapters/introduction.tex
work towards issue 16 - wip

There are many in-depth histories of the origin and development of writing
systems \citep[e.g.~][]{Robinson1995,Powell2012}, a story that we therefore will
not repeat here. However, the history of turning writing into computer readable
code is not so often told, so we decided to offer a short survey of the major
developments of such encoding here.\footnote{Because of the recent
history as summarized in this section, we have used mostly rather ephemeral
internet sources. When not referenced by traditional literature in the
bibliography, we have used \url{http://www.unicode.org/history/} and various
Wikipedia pages for the information presented here. A useful survey of the
historical development of the physical hardware of telegraphy and
telecommunication is \citet{Huurdeman2003}. Most books that discuss the
development of encoding of telegraphic communication focus of cryptography,
e.g.~\citet{Singh1999}, and forego the rather interesting story of ``open'',
i.e.~non-cryptographic, encoding that is related here.} This history turns out
to be intimately related to the history of telegraphic communication.

@@

Telegraphy was reinvigorated in the end of the eighteenth century through the
introduction of so-called semaphoric systems by Claude Chapelle to convey
messages over large distances. Originally, various specially designed
contraptions were used to send messages. Today, descendants of these systems are
still in limited use, for example utilizing flags or flashing lights. The
``innovation'' of those semaphoric systems was that all characters of the
written language were replaced one-to-one by visual signals. Since then, all
telegraphic systems have taken this principle, namely that any language to be
transmitted first has to be turned into some orthographic system, which
subsequently is encoded for transmission by the sender, and then turned back
into orthographic representation at the receiver side.\footnote{Sound and
video-based telecommunication of course takes a different approach by ignoring
the written version of language and directly encode sound waves or light
patterns.} This of course implies that the usefulness of any such telegraphic
encoding completely depends on the sometimes rather haphazard structure of
orthographic systems.

@@

From those bisignal encodings, true \textsc{binary codes} developed with a fixed
length of signals per character. In such systems only a single separator between
signals is needed, because the separation between characters can be established
by counting until a fixed number of signals has passed.\footnote{Of course, no
explicit separator is needed at all when the timing of the signals is known, which is
the principle used in all modern telecommunication systems. An important modern
consideration is also how to know where to start counting when you did not catch
the start of a message, something that is known in Unicode as \textsc{self
synchronization}.} In the context of electric telegraphy, such a binary code
system was first established by Émile Baudot in 1870, using a fixed combination
of five signals for each written character.\footnote{True binary codes have a
longer history, going at least back to the Baconian cipher devised by Francis
Bacon in 1605. However, the proposal by Baudot was the quintessential proposal
leading to all modern systems.} There are $2^5 = 32$ possible combination when
using five binary signals; an encoding today designated as 5-bit. These
codes are sufficient for all Latin letters, but of course they do not suffice
for all written symbols, including punctuation and digits. As a solution, the
Baudot code uses a so-called ``shift'' character, which signifies that from
that point onwards---until shifted back---a different encoding is used, allowing
for yet another set of 32 codes. In effect, this means that the Baudot code, and
the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an
extra bit of information, so the encoding is actually 6-bit (with $2^6
= 64$ different possible characters). For decades, this encoding was the
standard for all telegraphy and it is still in limited use today.

@@

Linguistically speaking, a \textsc{writing system} is a symbolic system that
uses visible or tactile signs to represent language in a systematic way. The
term writing system has two mutually exclusive meanings. First, it may
refer to the way a particular language is written. In this sense the term refers
to the writing system of a particular language, as, for example, in \emph{`the
Serbian writing system uses two scripts: Latin and Cyrillic.'} Second, the term
writing system may also refer to a type of symbolic system as used among the
world's languages to represent the language, as, for example, in
\emph{`alphabetic writing system.'} In this latter sense the term refers to how
scripts have been classified according to the way that they encode language, as
in, for example, \emph{`the Latin and Cyrillic scripts are both alphabetic
writing systems.'} To avoid confusion, this second notion of writing system
would more aptly have been called \textsc{script system}.

>>>>

There are many in-depth histories of the origin and development of writing
systems \citep[e.g.~][]{Robinson1995,Powell2012}, a story that we therefore will
not repeat here. However, the history of turning writing into computer readable
code is not so often told, so we decided to offer a short survey of the major
developments of such encoding here.\footnote{Because of the recent
history as summarized in this section, we have used mostly rather ephemeral
internet sources. When not referenced by traditional literature in the
bibliography, we have used \url{http://www.unicode.org/history/} and various
Wikipedia pages for the information presented here. A useful survey of the
historical development of the physical hardware of telegraphy and
telecommunication is \citet{Huurdeman2003}. Most books that discuss the
development of encoding of telegraphic communication focus of cryptography,
e.g.~\citet{Singh1999}, and forego the rather interesting story of open,
i.e.~non-cryptographic, encoding that is related here.} This history turns out
to be intimately related to the history of telegraphic communication.

@@

Telegraphy was reinvigorated in the end of the eighteenth century through the
introduction of so-called semaphoric systems by Claude Chapelle to convey
messages over large distances. Originally, various specially designed
contraptions were used to send messages. Today, descendants of these systems are
still in limited use, for example utilizing flags or flashing lights. The
innovation of those semaphoric systems was that all characters of the
written language were replaced one-to-one by visual signals. Since then, all
telegraphic systems have taken this principle, namely that any language to be
transmitted first has to be turned into some orthographic system, which
subsequently is encoded for transmission by the sender, and then turned back
into orthographic representation at the receiver side.\footnote{Sound and
video-based telecommunication of course takes a different approach by ignoring
the written version of language and directly encode sound waves or light
patterns.} This of course implies that the usefulness of any such telegraphic
encoding completely depends on the sometimes rather haphazard structure of
orthographic systems.

@@

From those bisignal encodings, true \textsc{binary codes} developed with a fixed
length of signals per character. In such systems only a single separator between
signals is needed, because the separation between characters can be established
by counting until a fixed number of signals has passed.\footnote{Of course, no
explicit separator is needed at all when the timing of the signals is known, which is
the principle used in all modern telecommunication systems. An important modern
consideration is also how to know where to start counting when you did not catch
the start of a message, something that is known in Unicode as \textsc{self
synchronization}.} In the context of electric telegraphy, such a binary code
system was first established by Émile Baudot in 1870, using a fixed combination
of five signals for each written character.\footnote{True binary codes have a
longer history, going at least back to the Baconian cipher devised by Francis
Bacon in 1605. However, the proposal by Baudot was the quintessential proposal
leading to all modern systems.} There are $2^5 = 32$ possible combination when
using five binary signals; an encoding today designated as 5-bit. These
codes are sufficient for all Latin letters, but of course they do not suffice
for all written symbols, including punctuation and digits. As a solution, the
Baudot code uses a so-called shift character, which signifies that from
that point onwards---until shifted back---a different encoding is used, allowing
for yet another set of 32 codes. In effect, this means that the Baudot code, and
the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an
extra bit of information, so the encoding is actually 6-bit (with $2^6
= 64$ different possible characters). For decades, this encoding was the
standard for all telegraphy and it is still in limited use today.

@@

Linguistically speaking, a \textsc{writing system} is a symbolic system that
uses visible or tactile signs to represent language in a systematic way. The
term writing system has two mutually exclusive meanings. First, it may
refer to the way a particular language is written. In this sense the term refers
to the writing system of a particular language, as, for example, in \emph{the
Serbian writing system uses two scripts: Latin and Cyrillic}. Second, the term
writing system may also refer to a type of symbolic system as used among the
world's languages to represent the language, as, for example, in
\emph{alphabetic writing system}. In this latter sense the term refers to how
scripts have been classified according to the way that they encode language, as
in, for example, \emph{the Latin and Cyrillic scripts are both alphabetic
writing systems}. To avoid confusion, this second notion of writing system
would more aptly have been called \textsc{script system}.

### 2017-12-26 book/chapters/ipa_meets_unicode.tex
work towards issue 16 - wip

As long as a transcription is only directed towards phonetically trained eyes,
then all the details of the Unicode-encoding are unimportant. For a linguist
reading an IPA transcription, many of the details that will be discussed in this
chapter might seem like hair-splitting trivialities. However, if IPA
transcriptions are intended to be used across resources (e.g.~searching similar
phenomena across different languages) then it becomes crucial that there are
strict encoding guidelines. Our main goal in this chapter is to present the
encoding issues and propose recommendations for a ``strict'' IPA encoding for
situations in which cross-resource consistency is crucial.

@@

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 12 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see~\cite[Appendix~C]{Moran2012}. The 159 characters from 12 
code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
as discussed in Section~\ref{ipa-recommendations}.}

@@

   \item The uses of the apostrophe has led to long discussions on the Unicode
        Standard email list. An English keyboard inputs the symbol
        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
        at \uni{02BC}, which unfortunately looks mostly extremely similar to
        \uni{2019}. 
  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Linguists are also unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard. For the ``real'' retroflex click, see 
        Section~\ref{ipa-additions}.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
        line} at \uni{007C}.
  \item For the marking of length there is a special Unicode characters, namely
        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.        
  \item There are two mostly identical-looking Unicode characters for the superscript
        <ʰ>: the \textsc{combining latin small letter h} at \uni{036A} and the
        \textsc{modifier letter small h} at \uni{02B0}. Making the situation 
        even more problematic is that they have different behavior (see 
        Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
        bevavior of <ʰ> with other superscript letters, we propose to 
        standardize on the modifier letter at \uni{02B0} (see 
        Section~\ref{pitfall-no-unique-diacritic-ordering}).

@@

   \item The dental click <ǀ> and the indication of a minor group break <|>
           look almost the same in
           most fonts. For a proper encoding, the \textsc{latin letter dental
           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
           should be used, respectively.
   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
           which according to the IPA is the character to by used for a major group 
           break (by intonation), to be encoded by \textsc{double vertical line} 
           at \uni{2016}.
   \item The marking of primary stress < ˈ > looks like an apostrophe, and
           is often typed with the same symbol as the ejective <\dia{02BC}>. For a
           proper encoding, these two symbols should be typed as 
           \textsc{modifier letter vertical line} at \uni{02C8} and
           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
   \item There are two different ``dashed''-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}. These of course
          look highly similar, although they are different sounds. As a solution, 
          we will actually propose to not use the middle tilde at all 
          (see Section~\ref{pitfall-multiple-options-ipa}).

@@

It is not just the Unicode Standard that offers multiple options for encoding
the IPA.\@ Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable when the goal is interoperability between resources written in
IPA.\@ We consider it crucial to distinguish between ``valid'' IPA, for which it
is sufficient that any phonetically-trained reader is able to understand the
transcription, and ``strict'' IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources. We are
aware of the following non-unique encoding options in the IPA, which will be
discussed in turn below:

@@

Most importantly though, in comparison to normal Unicode processing, the tie-bar
actually takes a reversed approach to complex graphemes. Basically, the Unicode
principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
sequences in a writing system have to be specified as ``tailored'' grapheme
clusters. Only in case the sequence is not a cluster, then this has to be
explicitly indicated. IPA takes a different approach. In IPA by default
different base letters are not connected into larger clusters; only when it is
specified in the string itself (using the tie bar).

@@

A similar situation arises with <\dia{0334}> \textsc{combining tilde overlay} at
\uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
the ``dark'' <l> in English codas, using the character <ɫ> \textsc{latin small
letter l with middle tilde} at \uni{026B}. This character is not canonically
equivalent to the combination <l>~+~<\dia{0334}>, so one of the two possible
encodings has to be chosen. Because the tilde overlay is described as a general
mechanism by the IPA, we propose to use the separated <\dia{0334}>
\textsc{combining tilde overlay} at \uni{0334}. However, note that phonetically 
this seems to be (almost) superfluous (see Section~\ref{pitfall-multiple-options-ipa}) 
and the typical usage in the form of <ɫ> is (almost) a homoglyph with <ɬ> (see 
Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we also suggest to try 
and avoid the tilde overlay completely.

@@

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode standard, which is probably best seen as an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive ``sucking'' sound similar
      to a fricative, for example <kǃ͓ʰ>; second, ``frictionalized'' vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ``ejective'' or ``creaky voicing'' are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

>>>>

As long as a transcription is only directed towards phonetically trained eyes,
then all the details of the Unicode-encoding are unimportant. For a linguist
reading an IPA transcription, many of the details that will be discussed in this
chapter might seem like hair-splitting trivialities. However, if IPA
transcriptions are intended to be used across resources (e.g.~searching similar
phenomena across different languages) then it becomes crucial that there are
strict encoding guidelines. Our main goal in this chapter is to present the
encoding issues and propose recommendations for a strict IPA encoding for
situations in which cross-resource consistency is crucial.

@@

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 12 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see~\cite[Appendix~C]{Moran2012}. The 159 characters from 12 
code blocks shown here are the characters proposed for strict IPA encoding, 
as discussed in Section~\ref{ipa-recommendations}.}

@@

   \item The uses of the apostrophe has led to long discussions on the Unicode
        Standard email list. An English keyboard inputs the symbol
        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
        at \uni{02BC}, which unfortunately looks mostly extremely similar to
        \uni{2019}. 
  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Linguists are also unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard. For the real retroflex click, see 
        Section~\ref{ipa-additions}.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
        line} at \uni{007C}.
  \item For the marking of length there is a special Unicode characters, namely
        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.        
  \item There are two mostly identical-looking Unicode characters for the superscript
        <ʰ>: the \textsc{combining latin small letter h} at \uni{036A} and the
        \textsc{modifier letter small h} at \uni{02B0}. Making the situation 
        even more problematic is that they have different behavior (see 
        Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
        bevavior of <ʰ> with other superscript letters, we propose to 
        standardize on the modifier letter at \uni{02B0} (see 
        Section~\ref{pitfall-no-unique-diacritic-ordering}).

@@

   \item The dental click <ǀ> and the indication of a minor group break <|>
           look almost the same in
           most fonts. For a proper encoding, the \textsc{latin letter dental
           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
           should be used, respectively.
   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
           which according to the IPA is the character to by used for a major group 
           break (by intonation), to be encoded by \textsc{double vertical line} 
           at \uni{2016}.
   \item The marking of primary stress < ˈ > looks like an apostrophe, and
           is often typed with the same symbol as the ejective <\dia{02BC}>. For a
           proper encoding, these two symbols should be typed as 
           \textsc{modifier letter vertical line} at \uni{02C8} and
           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
   \item There are two different dashed-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}. These of course
          look highly similar, although they are different sounds. As a solution, 
          we will actually propose to not use the middle tilde at all 
          (see Section~\ref{pitfall-multiple-options-ipa}).

@@

It is not just the Unicode Standard that offers multiple options for encoding
the IPA.\@ Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable when the goal is interoperability between resources written in
IPA.\@ We consider it crucial to distinguish between valid IPA, for which it
is sufficient that any phonetically-trained reader is able to understand the
transcription, and strict IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources. We are
aware of the following non-unique encoding options in the IPA, which will be
discussed in turn below:

@@

Most importantly though, in comparison to normal Unicode processing, the tie-bar
actually takes a reversed approach to complex graphemes. Basically, the Unicode
principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
sequences in a writing system have to be specified as tailored grapheme
clusters. Only in case the sequence is not a cluster, then this has to be
explicitly indicated. IPA takes a different approach. In IPA by default
different base letters are not connected into larger clusters; only when it is
specified in the string itself (using the tie bar).

@@

A similar situation arises with <\dia{0334}> \textsc{combining tilde overlay} at
\uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
the dark <l> in English codas, using the character <ɫ> \textsc{latin small
letter l with middle tilde} at \uni{026B}. This character is not canonically
equivalent to the combination <l>~+~<\dia{0334}>, so one of the two possible
encodings has to be chosen. Because the tilde overlay is described as a general
mechanism by the IPA, we propose to use the separated <\dia{0334}>
\textsc{combining tilde overlay} at \uni{0334}. However, note that phonetically 
this seems to be (almost) superfluous (see Section~\ref{pitfall-multiple-options-ipa}) 
and the typical usage in the form of <ɫ> is (almost) a homoglyph with <ɬ> (see 
Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we also suggest to try 
and avoid the tilde overlay completely.

@@

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode standard, which is probably best seen as an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive sucking sound similar
      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ejective or creaky voicing are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.
